Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2022-04-25 inst/surfing_jsa_rebuttal_referee2.tex

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{MS596 ``Analysis of competitive surfing tournaments
  with generalized Bradley-Terry likelihoods" by Driver and
  Hankin: second rebuttal}

Below, the reviewer's comments are in black, and our replies to the
issues are in \textcolor{blue}{blue}.  We have indicated changes to
the manuscript where appropriate.

Short story: we have accommodated all the comments with rewording and
clarification.

Slightly longer story: the referee makes a wide range of observations,
some good and some bad.  We have overhauled much of the writing, added
descriptions of the computational and mathematical methodology,
clarified ambiguous wording, produced new statistical tests, and
improved the visuals.

We have included an extended discussion of some of the points below.
The resulting document is, we believe, stronger and more scholarly
than before and we recommend it to you.

@@

\textcolor{blue}{These comments are entirely reasonable and, on
  re-reading the manuscript, we agree that we could have done a much
  better job describing the mathematical and statistical basis for our
  conclusions, especially the motivation for the likelihood equations.
  In the resubmission we have added a large amount of explanation that
  hopefully makes our reasoning more clear.\\ \\
  However, we observe that {\em all} our calculations are available
  for public inspection on gihub at our repo {\tt RobinHankin/hyper2}.
  In particular, file {\tt inst/surfing.Rmd} can reproduce all our
  findings.  However, the nitty-gritty detail of our R idiom is, we
  suggest, not really suitable for a JSA manuscript.  We believe that
  our enhanced reporting in the revision is a good compromise between
  not enough detail and too much.
  \begin{itemize}
  \item We have provided more detail (mathematical and computational)
    for the Fisher information argument for retaining the data with
    three surfers, as opposed to retaining just the pairs
  \item Added detail on the computational infrastructure for locating
    maximum likelihood estimates
  \item Referenced the computational method [{\tt equalp.test()} for
    testing the null of equal competitive abilities
  \item Introduced a new table that explicitly sets out the likelihood
    function for the reified entity Brazilian beachtype reified entity
  \end{itemize}
}

The figures and tables are poorly presented and would benefit from
significant rework in terms of formatting.

\textcolor{blue}{We have reworked the figures:
  \begin{itemize}
  \item replaced the pie chart with a dotchart (log scale)
  \item retouched the fine-scale detail of the scatterplot of points scored 
    against maximum likelihood estimates
  \item changed the caption of the ``Profile support for the Brazilian
    wavetype reified entity" diagram to be more informative.  We
    suspect that the low information density of this graph contributed
    to overall impression of the presentation alluded to by the
    comment above.  We would observe, however, that this particular
    diagram is worth retaining despite its plain appearance on two
    grounds: firstly, it supports our contention that the whole
    likelihood function is informative; and secondly, the smoothly
    varying curve is strong evidence that numerical optimization
    techniques are working correctly.
  \end{itemize}
  For the tables, we are happy to take specific advice, but in the
  meantime we have improved  some of the mathematical sections [using
    {\tt vphantom}] which reduces the cluttered appearance.
}

@@

From a style perspective, the paper is well-written and enjoyable to
read, which makes the lack of substantial content a bit of a shame, as
the authors are clearly able to write a paper worth reading. Overall,
it sounds like the authors do have the data and analysis available to
warrant a more complete paper, however, what they have submitted feels
rushed and rather rough around the edges. While possible, it would
take a fairly significant overhaul/expansion to the current version of
the paper to get it into a publication-ready format.

\textcolor{blue}{As discussed above, we have added new statistical
  tests and intepretations of existing tests.  We have rewritten and
  overhauled large sections of the writing.}

### 2022-04-25 inst/surfing.Rmd
appearance of the graphs

```{r showsurfingmaxppie, echo=FALSE, fig.height=8,fig.cap="dot chart of result"}
dotchart(surfing_maxp)
dotchart(log10(surfing_maxp))
pie(surfing_maxp)
pdf(file="showsurfingmaxppie.pdf")
pie(surfing_maxp)
dev.off()
jj <- log(surfing_maxp)
kk <- log(surfing_pairs_maxp)
names(jj) <- NULL
names(kk) <- NULL
l <- c(-4.6, -1.5)
dotchart(log(surfing_maxp),pch=NA,xlim=l)
mydotchart(jj,col="red" ,pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
mydotchart(kk,col="blue",pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
```

@@

pdf(file="directlogplotofstrengths.pdf")
plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength')
fit <- lm(log10(jjL)~log10(jjO))
for(i in seq_along(jjO)){
text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.5)
}
abline(fit)
dev.off()
```

@@

```{r plotwidep}
plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
abline(h= -2,lty=2)
pdf(file="plotwidep.pdf")
plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
abline(h= -2,lty=2)
dev.off()
```

>>>>

```{r showsurfingmaxppie, echo=FALSE, fig.height=8,fig.cap="dot chart of result"}
dotchart(surfing_maxp)
dotchart(log10(surfing_maxp))
pie(surfing_maxp)
pdf(file="showsurfingmaxppie.pdf")
pie(surfing_maxp)
dev.off()
jj <- log(surfing_maxp)
kk <- log(surfing_pairs_maxp)
names(jj) <- NULL
names(kk) <- NULL
l <- c(-4.6, -1.5)
dotchart(log(surfing_maxp),pch=NA,xlim=l)
mydotchart(jj,col="red" ,pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
mydotchart(kk,col="blue",pch=16,xlim=l,xaxt='n',frame.plot=FALSE)

pdf(file="showsurfingmaxpdotchart.pdf")
jj <- log10(surfing_maxp)
dotchart(jj,pch=16,xlab="Log of Bradley-Terry strength of competitors (base 10)")
dev.off()
```

@@

pdf(file="directlogplotofstrengths.pdf")
par(xpd=NA)
plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength',cex=1.5)
fit <- lm(log10(jjL)~log10(jjO))
for(i in seq_along(jjO)){
text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.8)
}
par(xpd=FALSE)
abline(fit)
dev.off()
```

@@

```{r plotwidep}
plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
abline(h= -2,lty=2)
pdf(file="plotwidep.pdf")
plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Bradley-Terry strength of the Brazilian beachtype reified entity",ylab="support")
abline(h= -2,lty=2)
dev.off()
```

### 2022-04-25 inst/surfing_jsa.tex
appearance of the graphs

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

@@

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots &WCT09&WCT10&WCT11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655\\
 $\ldots$   &            &   &   &       &       & \dots       &      &       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WCT
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{l|l}
order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
\end{tabular}
\caption{The \label{likelihoodcompatriot} six possible finishing
  orders (``$\succ$'' means ``scored more points than'') for the case
  where two compatriot surfers and an odd man out are in the water at
  the same time, with associated likelihood function for strengths
  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
  strength of the compatriot monster. and nationality and a third are
  in the water at the same time}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{l|l|l}
wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1)$\\ \hline
beach & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
point & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2}{p_1+p_2}$\\
\end{tabular}
\caption{Likelihood function for results between surfer 1 (Brazilian)
  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
\end{table}
\clearpage
\newpage

@@

\begin{figure}[h]
\includegraphics[width=4in]{showsurfingmaxppie.pdf} % Pie chart of strengths
\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
dominance of Ferreira, Florence, and Medina}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=4in]{directlogplotofstrengths.pdf}
\caption{Scatterplot of points scored  \label{compare_likelihood_points}
against maximum likelihood estimates}
\end{figure}

>>>>

Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

@@

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots  &WCT10&WCT11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
 $\vdots$   &            &   &   &       &       & $\vdots$&       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WCT
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{l|l}
order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+q+p_C}$\\
$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_2+q+p_C}$\\
$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_1+q+p_C}$\\
$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+p_2}$\\
$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_1+p_2}$\\
\end{tabular}
\caption{The \label{likelihoodcompatriot} six possible finishing
  orders (``$\succ$'' means ``scored more points than'') for the case
  where two compatriot surfers and an odd man out are in the water at
  the same time, with associated likelihood function for strengths
  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
  strength of the compatriot monster. and nationality and a third are
  in the water at the same time}
\end{table}

@@

\begin{table}[h]
\begin{tabular}{l|l|l}
wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1^{\vphantom{h}})$\\ \hline
beach & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
point & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2^{\vphantom{h}}}{p_1^{\vphantom{h}}+p_2}$\\
\end{tabular}
\caption{Likelihood function for results between surfer 1 (Brazilian)
  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
\end{table}
\clearpage
\newpage

@@

\begin{figure}[h]
\includegraphics[width=4in]{showsurfingmaxpdotchart.pdf} % pdf produced by surfing.Rmd
\caption{Dot chart\label{dotchartmax} showing maximum likelihood
  strengths of the 23 competitors in alphabetic order, logarithmic
  axis.  Note the dominance of Ferreira, Florence, and Medina}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}
\caption{Scatterplot of points scored  \label{compare_likelihood_points}
against maximum likelihood estimates}
\end{figure}
